Symbolic Dynamic Programming for Continuous State and Observation POMDPs

### 2012-11-08 trunk/cpomdp/camdp-xadd/doc/NIPS12/draft6.tex
NIPS Camera-ready

\newcommand{\xds}{\mathbf{dx}_s}
\newcommand{\xdsp}{\mathbf{dx}_s'}
\newcommand{\xdo}{\mathbf{dx}_o}
\newcommand{\open}{\mathit{open}}
\newcommand{\close}{\mathit{close}}
\newcommand{\high}{\mathit{high}}
\newcommand{\low}{\mathit{low}}
\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\renewcommand{\l}{\langle}
\renewcommand{\r}{\rangle}
\newcommand{\casemax}{\mathrm{casemax}}

@@

\section{DC-POMDP Model}

@@

%% TODO: further restrictions on observations!  also discrete problems!
%We assume familiarity with MDPs and
%We introduce discrete and continuous
%partially observable MDPs (DC-POMDPs) as an extension to 
%DC-MDPs~\cite{sanner_uai11}.  
A discrete and continuous
partially observable MDP (DC-POMDP) is a tuple $\langle
\mathcal{S},\mathcal{A},\mathcal{O},\mathcal{T},\mathcal{R},\mathcal{Z},\gamma,h
\rangle$.  States $\mathcal{S}$ are given by vector 
$\xds = (\vec{d}_s,\vec{x}_s) = (
d_{s_1},\ldots,d_{s_n},x_{s_1},\ldots,x_{s_m} )$ where each $d_{s_i}
\in \{ 0,1 \}$ ($1 \leq i \leq n$) is boolean and each
$x_{s_j} \in \mathbb{R}$ ($1 \leq j \leq m$) is continuous.
We assume a finite, discrete action space $\mathcal{A} = \{ a_1,
\ldots, a_r \}$. Observations
$\mathcal{O}$ are given by the vector $\xdo = (\vec{d}_o,\vec{x}_o) = (
d_{o_1},\ldots,d_{o_p},x_{o_1},\ldots,x_{o_q} )$ where each $d_{o_i}
\in \{ 0,1 \}$ ($1 \leq i \leq p$) is boolean and each $x_{o_j} \in
\mathbb{R}$ ($1 \leq j \leq q$) is continuous.

Three functions are required for modeling DC-POMDPs: (1) $\mathcal{T}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow  [ 0, 1 ]$ a Markovian transition model defined as the probability of the next state %$(\vec{d}_s',\vec{x}_s')$ 
given the action and previous state%
%$(\vec{d}_s,\vec{x}_s)$ 
%$p(\vec{d}',\vec{x}'|\cdots,a)$
; (2)  $\mathcal{R}:\mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R}$ a reward function which returns the immediate reward of taking an action in some state; and (3) an observation function defined as $\mathcal{Z} : \mathcal{S} \times \mathcal{A} \times \mathcal{O} \rightarrow [ 0, 1 ]$  which gives the probability of an observation given the outcome of a state after executing an action.  A discount factor $\gamma, \; 0 \leq \gamma \leq 1$ is used to discount rewards $t$ time steps into the future by $\gamma^t$.

We use a dynamic Bayes net (DBN)\footnote{We disallow general 
  synchronic arcs for simplicity of exposition but
  note their inclusion only places restrictions on the variable
  elimination ordering used during the dynamic programming backup
  operation.} to compactly represent the transition model $\mathcal{T}$ over the
factored state variables and we use a two-layer Bayes net to
represent the observation model $\mathcal{Z}$: {\footnotesize
\begin{align}
\mathcal{T}: \;\; &
%p(\vec{d}_s',\vec{x}_s'|\vec{d}_s,\vec{x}_s,a) = 
p(\xdsp|\xds,a) = 
\prod_{i=1}^n p(d_{s_i}'|\xds,a) \prod_{j=1}^m p(x_{s_j}'|\xds, \vec{d}_s',a). \label{eq:trans_model} \\
\mathcal{Z}: \;\; & 
%p(\vec{d}_o,\vec{x}_o|\vec{d}_s,\vec{x}_s,a) = 
p(\xdo|\xdsp,a) = 
\prod_{i=1}^p p(d_{o_i}|\xdsp,a) \prod_{j=1}^q p(x_{o_j}|\xdsp,a). \label{eq:obs_model}
\end{align}}
%% TRANSITION LINEAR, BELIEF, REWARD, OBSERVATION PWC???
Probabilities over \emph{discrete} variables $p(d_{s_i}'|\xds,a)$ and
$p(d_{o_i}|\xdsp,a)$ may condition on both discrete variables and
(nonlinear) inequalities of continuous variables; this is further
restricted to linear inequalities in the case of continuous
observations.  Transitions over \emph{continuous}
variables $p(x_{s_j}'|\xds, \vec{d}_s',a)$ must be deterministic (but
arbitrary nonlinear) piecewise functions;
%encoded using the Dirac $\delta$ function; 
in the case of continuous observations they are further restricted to
be piecewise linear; this permits discrete noise in the continuous 
transitions since they may condition on stochastically sampled
discrete next-state variables $\vec{d}_s'$.
%(hence allowing discrete noise, but not, e.g., Gaussian noise).
Observation probabilities over continuous variables $p(x_{o_j}|\xdsp,a)$ 
only occur in the case of continuous observation and are required to be
piecewise constant (a mixture of uniform distributions); the same
restriction holds for belief state representations.
The reward $R(\vec{d},\vec{x},a)$ may be 
an arbitrary (nonlinear) piecewise function in the case of
deterministic observations and a piecewise constant function in the
case of continuous observations.  
We now provide concrete examples.
%To make this concrete, we now provide examples of both discrete and
%continuous observation DC-POMDPs.

\textbf{Example} \textsc{\bf (Power Plant)~\cite{steam2}} \emph{The steam
generation system of a power plant evaporates feed-water under restricted 
pressure and temperature conditions to turn a stream turbine.
A reward is obtained when electricity is generated from the turbine 
and the steam pressure and temperature are within safe ranges.
Mixing water and steam makes the
respective pressure and temperature observations $p_o \in \mathbb{R}$
and $t_o \in \mathbb{R}$ on the underlying state $p_s \in \mathbb{R}$
and $t_s \in \mathbb{R}$ highly uncertain.  Actions $A = \{ \open, \close \}$
control temperature and pressure by means of a pressure valve.}

We initially present two DC-POMDP variants labeled \textsc{\bf 1D-Power
  Plant} using a single temperature state variable $t_s$.  
The transition and reward are common to both ---
temperature increments (decrements) with a closed (opened) valve, a
large negative reward is given for a closed valve with $t_s$ exceeding
critical threshold $15$, and positive reward is given for a safe, 
electricity-producing state:
% For the transition probability, we use the Dirac function to model the deterministic equations. We can also define stochasticity in the transition model using boolean random variables that are sampled stochastically.  The reward function can be any linear function of the state or action. Going above a threshold temperature (e.g. $T=10$) will cause an explosion in the plant when trying to close the valve which results in the negative reward of $-1000$. Staying below this temperature is safe and will produce electricity an gain the reward of $100$. The reward of an open valve is $-1$.
{\footnotesize
\vspace{-1mm}
\begin{align}
\label{eq:trans}
p(t_s'|t_s,a)= \delta\left[ t_s' - 
\begin{cases}
 (a=\open) &: t_s - 5 \\ 
(a = \close) &: t_s + 7 \\
\end{cases}
\right]
\hspace{2mm}
R(t_s,a) = 
\begin{cases}
 (a=\open) &: -1 \\
(a = \close)\wedge (t_s>15) &: -1000 \\
(a = \close)\wedge \neg(t_s>15) &: 100 \\
\end{cases} 
\end{align}
\vspace{-4mm}
}

@@

\section{Value Iteration for DC-POMDPs}

@@

In a DC-POMDP, the agent does not directly observe the states and thus
must maintain a belief state $b(\xds) = p(\xds)$.  For a
given belief state $\vec{b} = b(\xds)$, a POMDP policy $\pi$ can be
represented by a tree corresponding to a conditional plan $\beta$.  An
h-step conditional plan $\beta^h$ can be defined recursively in terms
of $(h-1)$-step conditional plans as shown in
Fig.~\ref{fig:cond_plan} (left).
Our goal is to find a policy $\pi$ that maximizes the value function,
defined as the sum of expected discounted rewards over horizon $h$
starting from initial belief state $\vec{b}$:
{\footnotesize
\vspace{-1mm}
\begin{equation}
V^h_\pi(\vec{b}) = E_{\pi} \left[ \sum\nolimits_{t=0}^{h} \gamma^t \cdot r_t \Big| \vec{b}_0 = \vec{b} \right]
\end{equation}
\vspace{-4mm}
}

@@

The $\Gamma^h$ in
this optimal $h$-stage-to-go value function can be computed via
Monahan's dynamic programming approach to \emph{value iteration}
(VI)~\cite{monahan82}.  Initializing $\alpha^0_1 = \vec{0}$, 
$\Gamma^0 = \{ \alpha^0_1 \}$, and assuming discrete 
observations $o \in \mathcal{O}^h$,
$\Gamma^h$ is obtained from
$\Gamma^{h-1}$ as follows:\footnote{The
  $\textrm{\large $\boxplus$}$ of sets is defined as $\textrm{\large
    $\boxplus$}_{j \in \{ 1,\ldots, n \} } S_j = S_1 \textrm{\large
    $\boxplus$} \cdots \textrm{\large $\boxplus$} S_n$ where the
  pairwise cross-sum $P \textrm{\large $\boxplus$} Q = \{ \vec{p} +
  \vec{q} | \vec{p} \in P, \vec{q} \in Q \}$.}  
{\footnotesize
\vspace{-1mm}
\begin{align} 
g^h_{a,o,j}(\xds) &=  \int_{\vec{x}_{s'}} \sum_{\vec{d_{s'}}} p(o|\xds',a)p(\xds'|\xds,a) \alpha^{h-1}_j(\xds') d\vec{x}_{s'}; \hspace{2mm}  \forall \alpha^{h-1}_{j} \in \Gamma^{h-1} \label{eq:backup} \\
\Gamma^{h}_a   &= R(\xds,a) + \gamma \textrm{\large $\boxplus$}_{o \in \mathcal{O}} \left\{ g^h_{a,o,j}(\xds) \right\}_j  \label{eq:cross_prod}\\ 
\Gamma^h  &= \bigcup_a \Gamma^h_a 
\end{align}
\vspace{-4mm}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\incmargin{.5em}
\linesnumbered
\begin{algorithm}[t!]
\footnotesize
\vspace{-.5mm}
\dontprintsemicolon
\SetKwFunction{backup}{Backup}
\SetKwFunction{genObs}{GenRelObs}
\SetKwFunction{prune}{Prune}
\SetKwFunction{remapWithPrimes}{Prime}
\Begin
{
   $V^0:=0, h:=0, \Gamma_{PBVI}^0 = \{ \alpha_1^0 \} $\;
   \While{$h < H$}
   {
       $h:=h+1, \Gamma^h :=\emptyset, \Gamma_{PBVI}^h :=\emptyset$\;
       \ForEach {$\vec{b}_i \in B$}
       {
       	\ForEach {$a \in A$}
      	 {
			$\Gamma_{a}^h :=\emptyset$ \;       		
       		\eIf {(continuous observations: $q > 0$)}
       	         {\emph{// Derive relevant observation partitions $\mathcal{O}_i^h$ for belief $\vec{b}_i$} \;
			  $\l \mathcal{O}_i^h,p(\mathcal{O}_i^h|\xdsp,a) \r \,:=\,$ \genObs{$\Gamma_{PBVI}^{h-1},a,\vec{b}_i$}\;}
                 {\emph{// Discrete observations and model already known}\;
                          $\mathcal{O}_i^h := \{ \vec{d}_o \}$; $p(\mathcal{O}_i^h|\xdsp,a) := $ see Eq~\eqref{eq:obs_model}\;}
       		 \ForEach {$o \in \mathcal{O}_i^{h}$}
       		 {
				\ForEach {$\alpha_j^{h-1} \in \Gamma_{PBVI}^{h-1}$}
       			{
   	 		  		$\alpha_j^{h-1} :=\,$ \remapWithPrimes{$\alpha_j^{h-1}$} 
   	 		  		\emph{// $\forall d_i$: $d_i \to d_i'$ and $\forall x_i$: $x_i \to x_i'$} \; 
%   	 		    	$\Gamma_{a,xd_{o_i},j}^h \,:=\, p(o|xd_{s}) \cdot$ \backup{$\alpha_j^{h},a$}\;
   	 		    	{$g_{a,o,j}^h \,:=\, $ see Eq~\eqref{eq:backup}}
       	      	}
%       	      	$\Gamma_{a,\xdo}^h\,:=\,\textrm{\large $\boxplus$} \Gamma_{a,xd_{o_i}}^h$\;
%       	      	$\Gamma_{a,\xdo}^h\,:=\, $ see Eq~\eqref{eq:cross_prod}\;
       	     }
%           $\Gamma_a^{h} \,:=\,R_a \oplus \gamma \cdot \Gamma_{a,\xdo}^h$\;
            $\Gamma_a^{h} \,:=\, $ see Eq~\eqref{eq:cross_prod}\;
            $\Gamma^{h} \,:=\, \Gamma^{h} \cup \Gamma_a^{h}$\;
       	 }
        	%monahan's pruning first generates all vectors, then prunes
                %but not for PBVI since we only retain alpha-functions provably dominant at a belief point
%              $\Gamma^h \,:=\, $\prune{$\Gamma^h$} \emph{// optional strict dominance testing of $\alpha$-functions}\; 
      }
             % $V^h \,:=\, \mathrm{max}_{\alpha_j \in \Gamma^h} \vec{b}_i \cdot \alpha_j$\;
             % $\pi^{*,h} \,:=\, \argmax_{a} \, \Gamma_a^{h}$\;
      \ForEach {$\vec{b}_i \in B$}
      {
     	$\alpha_{\vec{b}_i}^h :=\ \argmax_{\alpha_j \in \Gamma^h} \alpha_j \cdot \vec{b}_i$\;
     	$\Gamma_{PBVI}^h :=\ \Gamma_{PBVI}^h \cup \alpha_{\vec{b}_i}^h$\;
      }

       \If{$\Gamma_{PBVI}^h = \Gamma_{PBVI}^{h-1}$}
           {break $\,$ \emph{// Terminate if early convergence}\;}
   }
     \Return{$\Gamma_{PBVI}$} \;
}
\caption{\footnotesize \texttt{PBVI}(DC-POMDP, $H$, $B=\left\{\vec{b}_i \right\}$) $\longrightarrow$ $\l V^h \r$ \label{alg:vi}}
\vspace{-1mm}
\end{algorithm}
\decmargin{.5em}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Point-based value iteration (PBVI)} computes the value
function only for a set of belief states $\{ \vec{b}_i \}$ where
$\vec{b}_i := p(\xds)$.  The idea is straightforward and the main
modification needed to Monahan's VI approach in Algorithm~\ref{alg:vi}
(following~\cite{Perseus_cont}) is the loop from lines 22--24 where
only $\alpha$-vectors optimal at some belief state are retained for
subsequent iterations.  In the case of continuous observation
variables ($q > 0$), we will need to derive a relevant set of
observations on line 10, a key contribution of this work as described
in Section~\ref{sec:cont_obs}.  Whereas PBVI is optimal if all
reachable belief states within horizon $H$ are enumerated in $B$, in the DC-POMDP
setting, the generation of continuous observations will most often
lead to an infinite number of reachable belief states, even with
finite horizon.  Nonetheless, PBVI has been quite successful in
practice without exhaustive enumeration of all reachable
beliefs~\cite{pbvi_jair06,hsvi2,Perseus,gapmin}, which motivates our
use of PBVI in this work.

@@

and this is the form we use to represent all functions in a DC-POMDP.
The $\phi_i$ are disjoint logical formulae defined over $\xds$ and/or $\xdo$ with logical ($\land,\lor,\neg$) combinations of boolean variables and inequalities ($\geq,>,\leq,<$) over continuous variables.  
For discrete observation DC-POMDPs, the $f_i$ and inequalities may use any function (e.g., $\sin(x_1) > \log(x_2)\cdot x_3)$; for continuous observations, they are restricted to linear inequalities and linear or piecewise constant $f_i$ as described in Section~\ref{sec:model}.

@@

%xadd representation
The data structure of the \emph{extended algebraic decision diagram}
(XADD)~\cite{sanner_uai11} is used to support case statements and the
required operations.  Figure~\ref{fig:cond_plan} (right) 
is an example of an XADD representation.

>>>>

\newcommand{\xds}{\mathbf{x}_s,\!\mathbf{d}_s}
\newcommand{\xdsp}{\mathbf{x}_s',\!\mathbf{d}_s'}
%\newcommand{\xds}{\mathbf{dx}_s}
%\newcommand{\xdsp}{\mathbf{dx}_s'}
\newcommand{\xdo}{\mathbf{x}_o,\!\mathbf{d}_o}
%\newcommand{\xdo}{\mathbf{dx}_o}
\newcommand{\open}{\mathit{open}}
\newcommand{\close}{\mathit{close}}
\newcommand{\high}{\mathit{high}}
\newcommand{\low}{\mathit{low}}
\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\renewcommand{\l}{\langle}
\renewcommand{\r}{\rangle}
\newcommand{\casemax}{\mathrm{casemax}}

@@

\section{Hybrid POMDP Model}

@@

%% TODO: further restrictions on observations!  also discrete problems!
%We assume familiarity with MDPs and
%We introduce discrete and continuous
%partially observable MDPs (H-POMDPs) as an extension to 
%DC-MDPs~\cite{sanner_uai11}.  
A \emph{hybrid} (discrete and continuous)
\emph{partially observable MDP} (H-POMDP) is a tuple $\langle
\mathcal{S},\mathcal{A},\mathcal{O},\mathcal{T},\mathcal{R},\mathcal{Z},\gamma,h
\rangle$.  States $\mathcal{S}$ are given by vector 
$(\vec{d}_s,\vec{x}_s) = (
d_{s_1},\ldots,d_{s_n},x_{s_1},\ldots,x_{s_m} )$ where each $d_{s_i}
\in \{ 0,1 \}$ ($1 \leq i \leq n$) is boolean and each
$x_{s_j} \in \mathbb{R}$ ($1 \leq j \leq m$) is continuous.
We assume a finite, discrete action space $\mathcal{A} = \{ a_1,
\ldots, a_r \}$. Observations
$\mathcal{O}$ are given by the vector $(\vec{d}_o,\vec{x}_o) = (
d_{o_1},\ldots,d_{o_p},x_{o_1},\ldots,x_{o_q} )$ where each $d_{o_i}
\in \{ 0,1 \}$ ($1 \leq i \leq p$) is boolean and each $x_{o_j} \in
\mathbb{R}$ ($1 \leq j \leq q$) is continuous.

Three functions are required for modeling H-POMDPs: (1) $\mathcal{T}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow  [ 0, 1 ]$ a Markovian transition model defined as the probability of the next state %$(\vec{d}_s',\vec{x}_s')$ 
given the action and previous state%
%$(\vec{d}_s,\vec{x}_s)$ 
%$p(\vec{d}',\vec{x}'|\cdots,a)$
; (2)  $\mathcal{R}:\mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R}$ a reward function which returns the immediate reward of taking an action in some state; and (3) an observation function defined as $\mathcal{Z} : \mathcal{S} \times \mathcal{A} \times \mathcal{O} \rightarrow [ 0, 1 ]$  which gives the probability of an observation given the outcome of a state after executing an action.  A discount factor $\gamma, \; 0 \leq \gamma \leq 1$ is used to discount rewards $t$ time steps into the future by $\gamma^t$.

We use a dynamic Bayes net (DBN)\footnote{We disallow general 
  synchronic arcs for simplicity of exposition but
  note their inclusion only places restrictions on the variable
  elimination ordering used during the dynamic programming backup
  operation.} to compactly represent the transition model $\mathcal{T}$ over the
factored state variables and we use a two-layer Bayes net to
represent the observation model $\mathcal{Z}$: {\footnotesize
\begin{align}
\mathcal{T}: \;\; &
%p(\vec{d}_s',\vec{x}_s'|\vec{d}_s,\vec{x}_s,a) = 
p(\xdsp|\xds,a) = 
\prod_{i=1}^n p(d_{s_i}'|\xds,a) \prod_{j=1}^m p(x_{s_j}'|\xds, \vec{d}_s',a). \label{eq:trans_model} \\
\mathcal{Z}: \;\; & 
%p(\vec{d}_o,\vec{x}_o|\vec{d}_s,\vec{x}_s,a) = 
p(\xdo|\xdsp,a) = 
\prod_{i=1}^p p(d_{o_i}|\xdsp,a) \prod_{j=1}^q p(x_{o_j}|\xdsp,a). \label{eq:obs_model}
\end{align}}
%% TRANSITION LINEAR, BELIEF, REWARD, OBSERVATION PWC???
Probabilities over \emph{discrete} variables $p(d_{s_i}'\!|\xds,\!a)$ and
$p(d_{o_i}\!|\xdsp,\!a)$ may condition on both discrete variables and
(nonlinear) inequalities of continuous variables; this is further
restricted to linear inequalities in the case of continuous
observations.  Transitions over \emph{continuous}
variables $p(x_{s_j}'\!|\xds,\!\vec{d}_s',\!a)$ must be deterministic (but
arbitrary nonlinear) piecewise functions;
%encoded using the Dirac $\delta$ function; 
in the case of continuous observations they are further restricted to
be piecewise linear; this permits discrete noise in the continuous 
transitions since they may condition on stochastically sampled
discrete next-state variables $\vec{d}_s'$.
%(hence allowing discrete noise, but not, e.g., Gaussian noise).
Observation probabilities over continuous variables $p(x_{o_j}\!|\xdsp,\!a)$ 
only occur in the case of continuous observations and are required to be
piecewise constant (a mixture of uniform distributions); the same
restriction holds for belief state representations.
The reward $R(\vec{d},\vec{x},a)$ may be 
an arbitrary (nonlinear) piecewise function in the case of
deterministic observations and a piecewise constant function in the
case of continuous observations.  
We now provide concrete examples.
%To make this concrete, we now provide examples of both discrete and
%continuous observation H-POMDPs.

\textbf{Example} \textsc{\bf (Power Plant)~\cite{steam2}} \emph{The steam
generation system of a power plant evaporates feed-water under restricted 
pressure and temperature conditions to turn a steam turbine.
A reward is obtained when electricity is generated from the turbine 
and the steam pressure and temperature are within safe ranges.
Mixing water and steam makes the
respective pressure and temperature observations $p_o \in \mathbb{R}$
and $t_o \in \mathbb{R}$ on the underlying state $p_s \in \mathbb{R}$
and $t_s \in \mathbb{R}$ highly uncertain.  Actions $A = \{ \open, \close \}$
control temperature and pressure by means of a pressure valve.}

We initially present two H-POMDP variants labeled \textsc{\bf 1D-Power
  Plant} using a single temperature state variable $t_s$.  
The transition and reward are common to both ---
temperature increments (decrements) with a closed (opened) valve, a
large negative reward is given for a closed valve with $t_s$ exceeding
critical threshold $15$, and positive reward is given for a safe, 
electricity-producing state:
% For the transition probability, we use the Dirac function to model the deterministic equations. We can also define stochasticity in the transition model using boolean random variables that are sampled stochastically.  The reward function can be any linear function of the state or action. Going above a threshold temperature (e.g. $T=10$) will cause an explosion in the plant when trying to close the valve which results in the negative reward of $-1000$. Staying below this temperature is safe and will produce electricity an gain the reward of $100$. The reward of an open valve is $-1$.
{\footnotesize
\vspace{-1mm}
\begin{align}
\label{eq:trans}
p(t_s'|t_s,a)= \delta\left[ t_s' - 
\begin{cases}
 (a=\open) &: t_s - 5 \\ 
(a = \close) &: t_s + 7 \\
\end{cases}
\right]
\hspace{2mm}
R(t_s,a) = 
\begin{cases}
 (a=\open) &: -1 \\
(a = \close)\wedge (t_s>15) &: -1000 \\
(a = \close)\wedge \neg(t_s>15) &: 100 \\
\end{cases} 
\end{align}
\vspace{-4mm}
}

@@

\section{Value Iteration for Hybrid POMDPs}

@@

In an H-POMDP, the agent does not directly observe the states and thus
must maintain a belief state $b(\xds) = p(\xds)$.  For a
given belief state $\vec{b} = b(\xds)$, a POMDP policy $\pi$ can be
represented by a tree corresponding to a conditional plan $\beta$.  An
h-step conditional plan $\beta^h$ can be defined recursively in terms
of $(h-1)$-step conditional plans as shown in
Fig.~\ref{fig:cond_plan} (left).
Our goal is to find a policy $\pi$ that maximizes the value function,
defined as the sum of expected discounted rewards over horizon $h$
starting from initial belief state $\vec{b}$:
{\footnotesize
\vspace{-1mm}
\begin{equation}
V^h_\pi(\vec{b}) = E_{\pi} \left[ \sum\nolimits_{t=0}^{h} \gamma^t \cdot r_t \Big| \vec{b}_0 = \vec{b} \right]
\end{equation}
\vspace{-4mm}
}

@@

The $\Gamma^h$ in
this optimal $h$-stage-to-go value function can be computed via
Monahan's dynamic programming approach to \emph{value iteration}
(VI)~\cite{monahan82}.  Initializing $\alpha^0_1 = \vec{0}$, 
$\Gamma^0 = \{ \alpha^0_1 \}$, and assuming discrete 
observations $o \in \mathcal{O}^h$,
$\Gamma^h$ is obtained from
$\Gamma^{h-1}$ as follows:\footnote{The
  $\textrm{\large $\boxplus$}$ of sets is defined as $\textrm{\large
    $\boxplus$}_{j \in \{ 1,\ldots, n \} } S_j = S_1 \textrm{\large
    $\boxplus$} \cdots \textrm{\large $\boxplus$} S_n$ where the
  pairwise cross-sum $P \textrm{\large $\boxplus$} Q = \{ \vec{p} +
  \vec{q} | \vec{p} \in P, \vec{q} \in Q \}$.}  
{\footnotesize
\vspace{-1mm}
\begin{align} 
g^h_{a,o,j}(\xds) &=  \int_{\vec{x}_{s'}} \sum_{\vec{d_{s'}}} p(o|\xdsp,a)p(\xdsp|\xds,a) \alpha^{h-1}_j(\xdsp) d\vec{x}_{s'}; \hspace{2mm}  \forall \alpha^{h-1}_{j} \in \Gamma^{h-1} \label{eq:backup} \\
\Gamma^{h}_a   &= R(\xds,a) + \gamma \textrm{\large $\boxplus$}_{o \in \mathcal{O}} \left\{ g^h_{a,o,j}(\xds) \right\}_j  \label{eq:cross_prod}\\ 
\Gamma^h  &= \bigcup_a \Gamma^h_a 
\end{align}
\vspace{-4mm}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\incmargin{.5em}
\linesnumbered
\begin{algorithm}[t!]
\footnotesize
\vspace{-.5mm}
\dontprintsemicolon
\SetKwFunction{backup}{Backup}
\SetKwFunction{genObs}{GenRelObs}
\SetKwFunction{prune}{Prune}
\SetKwFunction{remapWithPrimes}{Prime}
\Begin
{
   $V^0:=0, h:=0, \Gamma_{PBVI}^0 = \{ \alpha_1^0 \} $\;
   \While{$h < H$}
   {
       $h:=h+1, \Gamma^h :=\emptyset, \Gamma_{PBVI}^h :=\emptyset$\;
       \ForEach {$\vec{b}_i \in B$}
       {
       	\ForEach {$a \in A$}
      	 {
			$\Gamma_{a}^h :=\emptyset$ \;       		
       		\eIf {(continuous observations: $q > 0$)}
       	         {\emph{// Derive relevant observation partitions $\mathcal{O}_i^h$ for belief $\vec{b}_i$} \;
			  $\l \mathcal{O}_i^h,p(\mathcal{O}_i^h|\xdsp,a) \r \,:=\,$ \genObs{$\Gamma_{PBVI}^{h-1},a,\vec{b}_i$}\;}
                 {\emph{// Discrete observations and model already known}\;
                          $\mathcal{O}_i^h := \{ \vec{d}_o \}$; $p(\mathcal{O}_i^h|\xdsp,a) := $ see Eq~\eqref{eq:obs_model}\;}
       		 \ForEach {$o \in \mathcal{O}_i^{h}$}
       		 {
				\ForEach {$\alpha_j^{h-1} \in \Gamma_{PBVI}^{h-1}$}
       			{
   	 		  		$\alpha_j^{h-1} :=\,$ \remapWithPrimes{$\alpha_j^{h-1}$} 
   	 		  		\emph{// $\forall d_i$: $d_i \to d_i'$ and $\forall x_i$: $x_i \to x_i'$} \; 
%   	 		    	$\Gamma_{a,xd_{o_i},j}^h \,:=\, p(o|xd_{s}) \cdot$ \backup{$\alpha_j^{h},a$}\;
   	 		    	{$g_{a,o,j}^h \,:=\, $ see Eq~\eqref{eq:backup}}
       	      	}
%       	      	$\Gamma_{a,\xdo}^h\,:=\,\textrm{\large $\boxplus$} \Gamma_{a,xd_{o_i}}^h$\;
%       	      	$\Gamma_{a,\xdo}^h\,:=\, $ see Eq~\eqref{eq:cross_prod}\;
       	     }
%           $\Gamma_a^{h} \,:=\,R_a \oplus \gamma \cdot \Gamma_{a,\xdo}^h$\;
            $\Gamma_a^{h} \,:=\, $ see Eq~\eqref{eq:cross_prod}\;
            $\Gamma^{h} \,:=\, \Gamma^{h} \cup \Gamma_a^{h}$\;
       	 }
        	%monahan's pruning first generates all vectors, then prunes
                %but not for PBVI since we only retain alpha-functions provably dominant at a belief point
%              $\Gamma^h \,:=\, $\prune{$\Gamma^h$} \emph{// optional strict dominance testing of $\alpha$-functions}\; 
      }
             % $V^h \,:=\, \mathrm{max}_{\alpha_j \in \Gamma^h} \vec{b}_i \cdot \alpha_j$\;
             % $\pi^{*,h} \,:=\, \argmax_{a} \, \Gamma_a^{h}$\;
      \emph{// Retain only $\alpha$-functions optimal at each belief point}\;
      \ForEach {$\vec{b}_i \in B$}
      {
     	$\alpha_{\vec{b}_i}^h :=\ \argmax_{\alpha_j \in \Gamma^h} \alpha_j \cdot \vec{b}_i$ $\,$\;
     	$\Gamma_{PBVI}^h :=\ \Gamma_{PBVI}^h \cup \alpha_{\vec{b}_i}^h$\;
      }

        \emph{// Terminate if early convergence}\;
       \If{$\Gamma_{PBVI}^h = \Gamma_{PBVI}^{h-1}$}
           {break $\,$\;}
   }
     \Return{$\Gamma_{PBVI}$} \;
}
\caption{\footnotesize \texttt{PBVI}(H-POMDP, $H$, $B=\left\{\vec{b}_i \right\}$) $\longrightarrow$ $\l V^h \r$ \label{alg:vi}}
\vspace{-1mm}
\end{algorithm}
\decmargin{.5em}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Point-based value iteration (PBVI)}~\cite{pbvi_jair06,Perseus} 
computes the value
function only for a set of belief states $\{ \vec{b}_i \}$ where
$\vec{b}_i := p(\xds)$.  The idea is straightforward and the main
modification needed to Monahan's VI approach in Algorithm~\ref{alg:vi}
is the loop from lines 23--25 where
only $\alpha$-functions optimal at some belief state are retained for
subsequent iterations.
In the case of continuous observation
variables ($q > 0$), we will need to derive a relevant set of
observations on line 10, a key contribution of this work as described
in Section~\ref{sec:cont_obs}.  Otherwise if the observations are only 
discrete ($q=0$), then a finite set of observations is already known and
the observation function as given in Eq~\eqref{eq:obs_model}.

We remark that Algorithm~\ref{alg:vi} is a generic framework that 
can be used for both PBVI and other variants of 
approximate VI.  If used for PBVI, an 
efficient direct backup computation of the optimal $\alpha$-function for belief 
state $\vec{b}_i$ should be used in line 17 
that is linear in the number of observations~\cite{pbvi_jair06,Perseus}
and which obviates the need for lines 23--25.  
However, for an alternate version of approximate value iteration that will often
produce more accurate values for belief states other than those in $B$, 
one may instead retain the full cross-sum backup
of line 17, but omit lines 23--25 --- 
this yields an approximate VI approach 
(using discretized observations 
relevant only to a chosen set of belief states $B$ if continuous observations
are present) that is not restricted to alpha-functions only optimal at $B$, 
hence allowing greater flexibility
in approximating the value function over all belief states.%}

Whereas PBVI is optimal if all reachable belief states within horizon
$H$ are enumerated in $B$, in the H-POMDP setting, the generation of
continuous observations will most often lead to an infinite number of
reachable belief states, even with finite horizon --- this makes
it quite difficult to provide optimality guarantees in the general case of PBVI
for continuous observation settings.  Nonetheless, PBVI has been quite
successful in practice without exhaustive enumeration of all reachable
beliefs~\cite{pbvi_jair06,hsvi2,Perseus,gapmin}, which motivates our
use of PBVI in this work.

@@

and this is the form we use to represent all functions in an H-POMDP.
The $\phi_i$ are disjoint logical formulae defined over $\xds$ and/or $\xdo$ with logical ($\land,\lor,\neg$) combinations of boolean variables and inequalities ($\geq,>,\leq,<$) over continuous variables.  
For discrete observation H-POMDPs, the $f_i$ and inequalities may use any function (e.g., $\sin(x_1) > \log(x_2)\cdot x_3)$; for continuous observations, they are restricted to linear inequalities and linear or piecewise constant $f_i$ as described in Section~\ref{sec:model}.

@@

%xadd representation
Case operations yield a combinatorial explosion in size if
na\"{i}vely implemented, hence 
we use the data structure of the \emph{extended algebraic
  decision diagram} (XADD)~\cite{sanner_uai11} as shown in
Figure~\ref{fig:cond_plan} (right) to \emph{compactly} 
represent case statements and \emph{efficiently} support the above
case operations with them.
